RCloud: Integrating Exploratory Visualization, Analysis and Deployment

### 2014-04-01 casestudies.tex
yes

To illustrate some aspects of RCloud, we present an example application
for stock price analysis.\footnote{Unfortunately this report could not
include production notebooks containing proprietary or personal information.}
The example, although simplified, shows key steps in the development
and deployment process we intend to support.

@@

Next, the notebook author or a collaborator may decide to extend
the application to also provide convenient access to the data,
apart from its visualization.
This is done by creating a notebook that defines a function.
That notebook then becomes a \emph{subroutine} for other notebooks.
It can be invoked from interactive notebooks, such as dashboards.
The data access notebook is is version-controlled like other notebooks.

Finally, consider the situation where an analyst wants to understand
the price dynamics of stocks with respects to various attributes and
time ranges. For this, interactive visualization may be very helpful.
In our example, the analyst creates an interactive tool with N views.

Through the tight integration of R and JavaScript functions and data,
the analyst writes a function in R which passes the dataframe and the
description of the charts to the dcplot library.  Simple R expressions
are captured as trees to generate JavaScript expressions.  The terse
chart description language, with intelligent defaults inspired by
ggplot2, provides a simple yet powerful interface to the grouping
and reduction functionality of crossfilter and the chart styles of
dc.js and d3.js.

>>>>

To illustrate some aspects of RCloud, we present an example application
for stock price analysis.\footnote{Unfortunately this report could not
include production notebooks containing proprietary or personal information.}
The example, although simplified, shows key steps in the development
and deployment process we support.

@@

Next, the notebook author or a collaborator may decide to extend
the application to also provide convenient access to the data,
apart from its visualization.
This is done by creating a notebook that defines a function.
That notebook then becomes a \emph{subroutine} for other notebooks.
It can be invoked from interactive notebooks, such as dashboards.
The data access notebook is version-controlled like other notebooks.

Finally, consider the situation where an analyst wants to understand
the price dynamics of stocks with respects to various attributes and
time ranges. For this, interactive visualization may be very helpful.
In our example, the analyst creates an interactive tool with multiple
linked views.

Through the tight integration of R and JavaScript functions and data,
the analyst writes a function in R which passes the dataframe and the
description of the charts to the dcplot library.  Simple R expressions
are captured as trees to generate JavaScript expressions.  The terse
chart description language, with intelligent defaults inspired by
ggplot2, provides a simple yet powerful interface to the grouping
and reduction functionality of the well-accepted charting libraries
crossfilter, dc.js and d3.js.

### 2014-04-01 discussion.tex
curation para
yes

@@

Asynchronous collaborative visual analytics
(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
\emph{of} the ACVA process. It will be important when we talk about
recommendation systems, and navigation of the set of notebooks, etc.

>>>>

{\bf Curation.} Even without more formal group management, users are
able to curate groups of related notebooks using the simple folder
system.  This is particularly effective for bringing together
training materials.

@@

%Asynchronous collaborative visual analytics
%(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
%\emph{of} the ACVA process. It will be important when we talk about
%recommendation systems, and navigation of the set of notebooks, etc.

### 2014-04-01 introduction.tex
yes

In addition, consider trends in interactive visualization software
in the open source community. Projects led by Jeff Heer
include Prefuse, Flare, Protovis, d3, and Vega, in that order.
All but the first of these libraries take advantage of web technologies,
and all but the second rely on open standards. Hadley Wickham is the
author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
static graphics, while ggvis provides convenient interactive visualizations
in R, using the R package Shiny and the Vega abstract visualization grammar,
opening a separate web server for each call. These projects point to
the trend toward visual analytics frameworks and systems that are
increasingly interactive, composable, and accessible through the web.

@@

\item Technology transfer. In our experience, development
of new analyses and visualizations is done by \emph{hackers}
and \emph{scripters} (using Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}) who are, by definition,
at least somewhat familiar with programming and software development.
On the other hand, \emph{application
users} ultimately derive value from the data by means of the tools
developed by hackers and scripters. Traditionally, the connection
between these two worlds is made by IT staff. In an environment where
business needs can change every day, manual IT support does not scale:
we need to create software infrastructure to connect the two worlds.
(The path backward is equally important: when problems discovered in
production require expert help, we would like the experts to be
able to immediately apply their full range of analytic tools without
moving the problem back to the EDA environment).
%% The two classes of users
%% have quite different perspectives but we
%% infrastructure for continuous deployment of visualizations.

\item Coexistence. In the current data science ecosystem, the
isolation of exploratory visualization and data analysis
environments is a hinderance to wider adoption of modern techniques
on both sides. We specifically see the richness of interactive
visualization tools as still being isolated from the richness
of statistical programming environments.
There is an opportunity to provide a framework so that developments
on either side are more easily employed by the other.
In addition, the integrated results should be more easily accessible
to the rest of the organization.

@@

RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
of statistical analysis tools. In fact, as we describe in
Section~\ref{sec:notebooks}, RCloud notebooks are sequences of R snippets
interspersed with Markdown for easy generation of reports. In
Section~\ref{sec:visualization}, we describe how RCloud enables
interactive visualization widgets in this setting. In
Section~\ref{sec:limitations}, we discuss some consequences of
designing for a single language, and how this was received by
our (small, but active) user community.

@@

The facilities for \emph{sharing, viewing and embedding} RCloud
notebooks address technology transfer, and are described in
Section~\todo{where?}. RCloud notebooks are edited and shared as web
pages. To support tight integration of R and HTML5, RCloud uses a
full bi-directional communication channel between the web browser
and a running R process, described in Section~\todo{where?}.

Because RCloud embraces technologies that are part of the computing
environment (specifically, inter-process communication via HTTP and
web services as remote procedure calls), we find it easier to
integrate RCloud with satellite efforts in corporate environments.
We discuss this architecture and consequences in Section~\ref{sec:system}.

The requirement of continuous technology transfer creates challenges
about the \emph{stability} of the many moving pieces: if notebooks are
updated and deployed automatically, how can the system handle
changing requirements and public interfaces? We discuss our choice of
transparent, integrated version control for RCloud notebooks in 
Section~\ref{sec:notebooks}.

>>>>

In addition, consider trends in interactive visualization software
in the open source community.  Projects led by Jeff Heer
include Prefuse, Flare, Protovis, d3, and Vega, in that order.
All but the first of these libraries take advantage of web technologies,
and all but the second rely on open standards. Hadley Wickham is the
author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
static graphics, while ggvis provides convenient interactive visualizations
in R, using the R package Shiny and the Vega abstract visualization grammar,
opening a separate web server for each call. These projects show
a trend toward visual analytics frameworks and systems that are
increasingly rich, interactive, composable, and accessible through the web.

@@

\item Technology transfer. In our experience, development
of new analyses and visualizations is done by \emph{hackers}
and \emph{scripters} (using Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}) who are, by definition,
at least somewhat familiar with programming and software development.
On the other hand, \emph{application
users} ultimately derive value from the data by means of the tools
developed by hackers and scripters. Traditionally, the connection
between these two worlds is made by IT staff. In an environment where
business needs can change every day, manual IT support does not scale:
we need to create software infrastructure to connect the two worlds.
%% The two classes of users
%% have quite different perspectives but we
%% infrastructure for continuous deployment of visualizations.

\item Coexistence. In the current data science ecosystem, the
isolation of exploratory visualization and data analysis
environments is a hinderance to wider adoption of modern techniques
on both sides. We specifically see the richness of interactive
visualization tools as still being isolated from the richness
of statistical programming environments.
(The path backward from production to analysis can be 
equally important: when problems discovered in
production require expert help, we would like experts to be
able to immediately apply the full range of analytic tools without
extracting data to move it back to the EDA environment.)
There is an opportunity to provide a framework so that developments
on either side are more easily employed by the other.
In addition, the integrated results should be more easily accessible
to the rest of the organization.

@@

RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
of statistical analysis tools. As described in Section~\ref{sec:notebooks},
RCloud notebooks are sequences of R snippets interspersed with Markdown
for easy generation of reports. We also describe how RCloud enables
interactive visualization widgets in this setting (Section~\ref{sec:executinginbrowser}).

@@

The facilities for \emph{sharing, viewing and embedding} RCloud
notebooks address technology transfer.
Because RCloud embraces technologies that are part of the computing
environment (specifically, inter-process communication via HTTP and
web services as remote procedure calls), it is not complicated to
integrate RCloud with satellite efforts in corporate environments.

The requirement of continuous technology transfer, though, creates challenges
about the \emph{stability} of the many moving pieces: if notebooks are
updated and deployed automatically, how can the system handle
changing requirements and public interfaces? We discuss our choice of
transparent, integrated version control for RCloud notebooks in 
Section~\ref{sec:deployment}.

In Section~\ref{sec:limitations}, we discuss some consequences of
designing for a single language, and how this was received by
our (small, but active) user community.

### 2014-04-01 relatedwork.tex
yes

Kandel et al. argue in their review of data wrangling work that data
cleaning, wrangling and transformation is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
does not by itself include modules specific to data cleaning and
wrangling, we note from internal experience that these cleaning
scripts themselves tend to change over time. RCloud addresses this
issue by providing easy publishing of data-cleaning notebooks as web
services, which reduces the total data-cleaning effort across an
organization.

Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this impedance mismatch. They also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is a difficult and important problem, and that sharing
and versioning of data sources and artifacts is hindered by current
technology in practice. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their work points to the opportunity for better technology
to support collaboration and sharing by data analysis teams.

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses cost of
integration, content export, presentation and view sharing. Notebooks,
and the integrated version control system for them, described in
Section~\ref{sec:notebooks}, address modularity and granularity, and
artifact histories.

@@

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer data-analysis to analysts accustomed to creating and
executing R and Python code while keeping many of the attractive
properties like automatic versioning and management.

@@

One overall goal is to reduce the gap between implementers and deployers
of technology in visual analytics. The fusion of development with production
operations in software release management ("devops" \cite{Httermann:2012:DD})
is a trend in web services and related fields. By making it possible
for data scientists to expend a little more effort in the process of
creating experiments, it may be possible to eliminate the need for
programming teams to recreate their results in production with
a concomitant penalty in time and expense.

>>>>

Kandel et al. argue in their review of data wrangling work that data
cleaning, wrangling and transformation is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
does not by itself include modules specific to data cleaning and
wrangling, we note from practical experience that these cleaning
scripts themselves tend to change over time. RCloud addresses this
issue by providing easy publishing of data-cleaning notebooks as web
services, which reduces the total data-cleaning effort across an
organization.

Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this mismatch. They also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is a difficult and important problem, and that sharing
and versioning of data sources and artifacts is hindered by current
technology in practice. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their work points to the opportunity for better technology
to support collaboration and sharing by data analysis teams.

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF}.
Notebooks, and the integrated version control system for them, described in
Section~\ref{sec:notebooks}, address modularity and granularity, and
artifact histories.
\emph{Starring}, the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses cost of
integration, content export, presentation and view sharing.

@@

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer match for analysts accustomed to creating and
executing R and Python code, while retaining many attractive
properties like automatic versioning and management.

@@

One overall goal is to reduce the gap between implementers and deployers
of technology in visual analytics. The fusion of development with production
operations in software release management (``DevOps'' \cite{Httermann:2012:DD})
is a trend in web services and related fields. By making it possible
for data scientists to expend a little more effort in the process of
creating experiments, it may be possible to eliminate the need for
programming teams to recreate their results in production with
a concomitant penalty in time and expense.

### 2014-04-01 system.tex
yes

The central of the screen is for editing code and visualizations through
the notebook interface. Supplementary ``asset'' files which are part of the
notebook but not part of the notebook flow, are edited on the right.
Controls at the top of the screen allow the user to quickly run, view, 
and share the notebook. The call mechanism is the URL itself; arguments are
read from the URL and the final cell is the result. On the left are the 
browsing area for seeing other users' notebooks and for searching, as well
as the help system.

@@

Using GitHub was mainly a tactical decision to reduce our
time-to-prototype. At the same time, it had the beneficial effect of
allowing our backend to communicate with separate processes over HTTP
and JSON. This had the serendipitous advantage of quickly enabling
full search over RCloud notebooks, using Apache SOLR.
%
This perspective has implications for
systems-based research and prototyping in visual analytics:
For small teams to efficiently explore novel ideas toward the
improvement of a large, complex software ecosystem, it pays off
to use existing solutions as far as possible.
%
Even though it involved learning standards that are not directly
related to visual analytics, our early decision to use the GitHub
gist API was successful. Future projects can gain many
of the same benefits by adopting a similar strategy.

@@

Implicit signaling of interest is supported by keeping click-through
\cite{Joachims:2005:AIC} and execution counts. (In addition to these
classic techniques of collecting feedback from web search, we anticipate
applying static and dynamic code analysis to generate fine-grained
information about relationships, for example, which packages and data
sets often appear together.)

\subsection{Executing R in a web browser}

One our main goals is to provide ubiquitous access to the statistical
tools of the R programming language in the wider environment in which data
science teams participate.
%
Given the recent developments of interactive visualization and visual
analytics in HTML5 and the web, we decided to create an
\emph{interconnect} between the two environments.
%
Every RCloud session spawns a new R execution process in a remote
server. The communication between the web browser and the

@@

HTTP and web services are convenient but much of the protocol is
stateless. (For example, GET requests are required to not change the
remote state, and results can be cached in transparent proxies along
the network). In the case of close communication between a running
R process and a web browser, we need something that is more flexible
than distinct URLs for every R process, and which embraces
statefulness of both calculations and insights in progress.

\subsection{Interactive notebooks\label{sec:notebooks}}

@@

Because notebooks are continually published, any work is immediately
available both as a subroutine and visual component. Close colleagues
can start on the next stage of analysis, or delve into the data,
even while the original author is polishing the algorithm or its
presentation. The code is the page, and can either be shaped into
a function of inputs and outputs, or the linear cells of the notebook
can be reworked into a full-fledged HTML layout.

@@

>>>>

The center panel of the screen is for editing code and visualizations through
the notebook interface. Supplementary ``asset'' files which are part of the
notebook but not part of the notebook flow, are edited on the right.
Controls at the top of the screen allow the user to quickly run, view, 
and share the notebook. The call mechanism is the URL itself; arguments are
read from the URL and the final cell is the result. On the left are the 
browsing area for seeing other users' notebooks and for searching, as well
as a help system.

@@

Using GitHub was mainly a tactical decision to reduce our
time-to-prototype. At the same time, it had the beneficial effect of
allowing our backend to communicate with separate processes over HTTP
and JSON. This had the serendipitous advantage of quickly enabling
full search over RCloud notebooks, using Apache SOLR.
%
This perspective has implications for
systems-based research and prototyping in visual analytics:
for small teams to efficiently explore novel ideas toward the
improvement of a large, complex software ecosystem, it pays
to use existing solutions as far as possible.
%
Even though it involved learning standards that are not directly
related to visual analytics, our early decision to use the GitHub
gist API was successful. Future projects can gain many
of the same benefits by adopting a similar strategy.

@@

Implicit signaling of interest is supported by keeping click-through
\cite{Joachims:2005:AIC} and execution counts. (In addition to these
classic techniques of collecting feedback from web search, we anticipate
applying static and dynamic code analysis to infer fine-grained
information about relationships, for example, which packages and data
sets often appear together.)

\subsection{Executing R in a web browser\label{sec:executinginbrowser}}

One our main goals is to provide ubiquitous access to the statistical
tools of the R programming language in the wider environment in which data
science teams participate.
%
Given the recent developments of interactive visualization and visual
analytics in HTML5 and the web, we decided to create an
\emph{interconnect} between the two environments.
%
Every RCloud session spawns a new R execution process in a remote
server.

@@

HTTP and web services are convenient but much of the protocol is
stateless. (For example, GET requests are required to not change the
remote state, and results can be cached in transparent proxies along
the network). In the case of close communication between a running
R process and a web browser, we need something that is more flexible
than distinct URLs for every R process, and which embraces
statefulness of both calculations and visualization parameters.

\subsection{Interactive notebooks\label{sec:interactivenotebooks}}

@@

@@

Because notebooks are continually published, any work is immediately
available both as a subroutine and a visual component. Close colleagues
can start on the next stage of analysis, or delve into the data,
even while the original author is polishing an algorithm or its
presentation. The code is the page, and can either be shaped into
a function of inputs and outputs, or the linear cells of the notebook
can be reworked into a full-fledged HTML layout.
